Alternative Methods for Implementing Explicit and Finding Implicit Sharing in embedded DSLs

### 2022-12-15 paper/paper.tex
just missing abstract and memory profiling

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
\usepackage{listings}
\usepackage{minted}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
\lstset{language=haskell}
%
\title{Alternative Methods for Implementing
    Explicit and Finding Implicit Sharing in embedded DSLs}
%
\titlerunning{Alternative Explicit and Implicit Sharing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Curtis D'Alves \and
Christopher Anand \and
Lucas Dutton \and
Steven Gonder
}
%
\authorrunning{C. D'Alves et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{McMaster University, 1280 Main St W Hamilton, Canada}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
TODO The abstract should briefly summarize the contents of the paper in
150--250 words.

@@

TODO describe sharing problem (mention observable sharing
\cite{gill:observablesharing} and implicit/explicit sharing
\cite{kiselyov:sharing} papers)

TODO describe finally tagless \cite{carette:finallytagless}

We present methods for implementing embedded DSLs with sharing that are both
safe and maintain all the benefits of being embedded in the Haskell ecosystem.
This means DSL functions are type-safe, do not require the use of unsafe
referencing (i.e., via unsafePerformIO) and can return Haskell's container
types (i.e., tuples, lists, etc) without breaking sharing.

\section{Detecting Sharing}

@@

-- Example
v0 = Variable ``v0''
exp0 = Add v0 (Constant 0)
exp1 = Add exp0 exp0
\end{minted}

@@

TODO describe DAG conversion by pointer comparison in the state monad

\subsection{Detecting Sharing In Finally Tagless DSLs}

@@

We can then create different instances to implement different functionality.
For example, we can generate the AST from the previous DSL like so
\begin{minted}{haskell}
newtype ExpR a = ExpR { unExpR :: Exp }

instance Exp ExpR where
  constant = ExpR . Constant
  variable = ExpR . Variable
  add (ExpR x) (ExpR y) = ExpR (Add x y)
\end{minted}

Or we can implement pretty printing
\begin{minted}{haskell}
newtype Pretty a = Pretty { unPretty :: String }

@@

And use the same DSL code to run either implementation
\begin{minted}{haskell}
exp :: Exp repr => repr Int -> repr Int
exp v0 =
  let
    exp0 = add v0 (constant 0)
  in add exp0 exp0

expR = unExpR $ exp $ variable "v0"
expP = unPretty $ exp $ variable "v0"
\end{minted}

@@

\subsection{Implicit Sharing Via Hash-Consing}
TODO
cite Ershov’s original description of hash-consing \cite{ershov1958:consing}
cite Type safe consing implementation (with performance benchmarks) \cite{filliatre:typesafeconsing}

In \cite{kiselyov:sharing}, a solution for detection implicit sharing in finally
tagless style is presented via the method of hash-consing. You can find a more
throrough explanation of the method there, but we'll give an overview here. This
method first involves defining a DAG type, for example
\begin{minted}{haskell}
type NodeID = Int
data Node = NAdd NodeID NodeID
          | NVariable String
          | NConstant Int

@@

Note the purpose of the BiMap type is to be able to quickly insert and lookup
nodes by their NodeID (i.e., a bijection of Node's and their NodeID's), and is
most optimally implemented as a hash table with linear probing. The
representation for the finally tagless instance is then a wrapper around a State
monad that holds DAG in its state and returns the current (top) NodeID.

@@

The trick to uncovering sharing in the implementation is implemented via the
{\bf hashcons} function, which inserts a new node into the current DAG, but not
before checking if it is already there.
\begin{minted}{haskell}
hashcons :: Node -> State DAG NodeID
hashcons e = do
  DAG m <- get
  case lookup_key e m of
    Nothing -> let (k,m') = insert e m
               in put (DAG m') >> return k
    Just k -> return k
\end{minted}

\subsection{Limitations of Hash-Consing}

@@

Implicit sharing via hash-consing prevents duplication in the resulting DAG, but
unfortunately doesn't prevent redundant computation. Consider the following
equivalent attempt at using Haskell's built-in sharing in the finally tagless DSL
\begin{minted}{haskell}
dslSharing :: Exp Graph -> Exp Graph -> Exp Graph
dslSharing x y =
  let
    z = add x y
  in add z z
\end{minted}
Note {\bf z} is a wrapper around a State monad. Recall the implementation of
{\bf add} via hash consing
\begin{minted}{haskell}
add e1 e2 = Graph (do
                   h1 <- unGraph e1
                   h2 <- unGraph e2
                   hashcons $ NAdd h1 h2)
\end{minted}
The values {\bf h1} and {\bf h2} need to be explicitely evaluated through the
State monad, meaning even if {\bf e1} and {\bf e2} are the same shared Haskell
value, their underlying computations will be performed twice. Hash-consing will
prevent these redundencies from appearing in the resulting DAG, but the entire
unshared AST wil still be traversed, performing a hash-cons on each node.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/hashcons.png}
  \caption{Number of hashcons to add operations performed} \label{fig:hashcons}
\end{figure}
As you can see from figure \ref{fig:hashcons}, this code will perform approximately
$2^{n+1}$ hashcons operations, where n is the number of {\bf add}'s.

\subsection{Explicit Sharing and Limitations}

@@

addChains x =
  let_ x (\x0 ->
  let_ (add x0 x0)  (\x1 ->
  let_ (add x1 x1)  (\x2 ->
   ...
  )))
\end{minted}
This makes the code a bit clunky and adds an extra burden on the DSL writer, but
it prevents unnecessary hash-consing in our example. However the method has it's
limitations, suppose we want to write a DSL function that returns multiple
outputs, such as tuples or container types like lists (for example
\mintinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr Int)}). %TODO
                                % need to give implementation of this function?
First of all, we need to implement different versions of the custom let construct to
correspond to the number of outputs
\begin{minted}{haskell}
class ExpLet repr where
  let_ :: repr a -> (repr a -> repr b) -> repr b
  let_ 2 :: repr a -> (repr a -> (repr b,repr c)) -> (repr b,repr c)

instance ExpLet repr where
   ...
   let_2 e f = let
       g0 = Graph (do x <- unGraph e
                        let (o0,o1) = f (Graph (return x))
                        unGraph o0)
       g1 = Graph (do x <- unGraph e
                        let (o0,o1) = f (Graph (return x))
                        unGraph o1)
       in (g0,g1)
\end{minted}
If need be, its possible to enumerate custom let instances for every amount of
outputs or container types we would need, we could even use template Haskell to
accomplish this. However, this custom let construct now has a new source of
redundancy in its outputs. Each output it returns will now have to individually
evaluate it's input, so a chain of DSL functions that output 2 or more values
will suffer from the same exponential scaling of hashcons.

@@

instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (hashcons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (hashcons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST (NAdd undefined undefined) [unStringAST e1,unStringAST e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> hashcons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST

@@

-- TODO better example?
test x y = let
  z = cache "z" (add x y)
  in add z z
\end{minted}

The cache operation replaces the current byte string AST with a new label, and
we'll define a new operation \mintinline{haskell}{runCache} that will check if
the label already exists in the cache map before inserting it.
\begin{minted}{haskell}
runCache :: ByteString -> Maybe ByteString -> Trie ByteString -> State DAG ()
runCache sAST mAddCache cacheMap = do
  case mAddCache of
    Nothing -> return ()
    Just sAST0 ->
      case Trie.lookup sAST cacheMap of
        Nothing -> let cacheMap' = Trie.insert sAST sAST0 cacheMap
                   in modify (\dag -> dag { dagCache = cacheMap' })
        Just sAST1 -> if sAST1 == sAST0
                         then return ()
                         else error $ "attempted to recache: " ++ show sAST

seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST mCache) =
      do DAG trie cacheMap _ <- get
         runCache sAST mCache cacheMap
         ...
\end{minted}
We need to make sure we don't attempt to insert the same cache label for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the State monad to prevent or modify the subtitution. The best we can do is
crash the program, or if we use a monad tansformer, we ould use
Control.Monad.Except to through an exception. Either way it's up to the DSL
writer to insure they don't reuse the same label.

\section{Conclusion}
TODO

\subsubsection{Acknowledgements} Please place your acknowledgments at
the end of the paper, preceded by an unnumbered run-in heading (i.e.
3rd-level heading).

@@

% LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless Trie
% Local Variables:
% LaTeX-verbatim-environments-local: ("minted")
% eval: (setq-local LaTeX-indent-environment-list (cons '("minted" current-indentation) (default-value 'LaTeX-indent-environment-list)))
% End:

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
\usepackage{listings}
\usepackage{minted}

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
\lstset{language=haskell}
%
\title{Alternative Methods for Implementing
    Explicit and Finding Implicit Sharing in embedded DSLs}
%
\titlerunning{Alternative Explicit and Implicit Sharing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Curtis D'Alves \and
Christopher Anand \and
Lucas Dutton \and
Steven Gonder
}
%
\authorrunning{C. D'Alves et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{McMaster University, 1280 Main St W Hamilton, Canada}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
TODO The abstract should briefly summarize the contents of the paper in
150--250 words.

@@

In his paper ``Implementing explicit and finding implicit sharing in embedded
DSLs'' \cite{kiselyov:sharing}, Kiselyov presents a method for implementing
eDSLs in finally tagless form that generate a directed acyclic graph (DAG) with
sharing. The interface for such DSL's uses the following format
\begin{minted}{haskell}
class Exp repr where
  variable :: String -> repr Int
  constant :: String -> repr Int
  add :: repr Int -> repr Int -> repr Int
  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
However, as we will explain in sections~\ref{limithashcons} and
\ref{limitexplicit}, when the DSL contains functions which return multiple
outputs (e.g., tuples, lists, etc.) such as \mintinline{haskell}{novel} in the
above code snippet, Kiselyov's method of detecting sharing scales exponentially,
and their method of explicitly declaring sharing is inapplicable. In our work,
this translated into the inability to process large library functions.

In this paper, we will review Kiselyov's methods, identifying core issue, and
present methods for implementing embedded DSLs with sharing that are both safe
and maintain all the benefits of being embedded in the Haskell ecosystem. This
means DSL functions are type-safe, do not require the use of unsafe referencing
(i.e., via unsafePerformIO) and can return Haskell's container types (i.e.,
tuples, lists, etc) without breaking sharing.

\section{Background: Detecting Sharing}

@@

-- Example
v0 = Variable "v0"
exp0 = Add v0 (Constant 0)
exp1 = Add exp0 exp0
\end{minted}

@@

TODO describe sharing through pointer comparison in monads
\subsection{Finally Tagless DSLs}

@@

We can then create different instances to implement different functionality.
For example, we can implement a pretty printer for our AST like so
\begin{minted}{haskell}
newtype Pretty a = Pretty { unPretty :: String }

@@

@@

\subsection{Implicit Sharing Via Hash-Consing}

In this section, we'll provide an overview of Kiselyov's method in
\cite{kiselyov:sharing} for detection of implicit sharing in finally tagless
style is presented via the method of hash-consing, and discuss it's limitations
in more detail. This method first involves defining a DAG type, for example
\begin{minted}{haskell}
type NodeID = Int
data Node = NAdd NodeID NodeID
          | NVariable String
          | NConstant Int

@@

Note the purpose of the BiMap type is to be able to quickly insert and lookup
nodes (i.e., a bijection of node's and their unique identifiers), and is most
optimally implemented as a hash table with linear probing. Such an
implementation of hash-consing using linear probing for an embedded DSL is given
in thesis \cite{thai2021type}. The representation for the finally tagless
instance is then a wrapper around a State monad that holds DAG in its state and
returns the current (top) NodeID.

@@

The trick to uncovering sharing in the implementation is implemented via the
{\bf hashcons} function, which inserts a new node into the current DAG, but not
before checking if it is already there.
\begin{minted}{haskell}
hashcons :: Node -> State DAG NodeID
hashcons e = do
  DAG m <- get
  case lookup_key e m of
    Nothing -> let (k,m') = insert e m
               in put (DAG m') >> return k
    Just k -> return k
\end{minted}
The technique is essentially that of
hash-consing, popularized by its use in LISP compilers, but discovered long
before that in 1958 \cite{ershov1958:consing}. Other works have explored the use
of type safe hash consing in embedded DSLs, see \cite{filliatre:typesafeconsing}.

\subsection{Limitations of Hash-Consing} \label{limithashcons}

@@

Implicit sharing via hash-consing prevents duplication in the resulting DAG, but
unfortunately doesn't prevent redundant computation. Consider the following
equivalent attempt at using Haskell's built-in sharing in the finally tagless DSL
\begin{minted}{haskell}
dslSharing :: Exp Graph -> Exp Graph -> Exp Graph
dslSharing x y =
  let
    z = add x y
  in add z z
\end{minted}
Note {\bf z} is a wrapper around a State monad. Recall the implementation of
{\bf add} via hash consing, the values {\bf h1} and {\bf h2} need to be
explicitely evaluated through the State monad, meaning even if {\bf e1} and {\bf
  e2} are the same shared Haskell value, their underlying computations will be
performed twice. Hash-consing will prevent these redundencies from appearing in
the resulting DAG, but the entire unshared AST wil still be traversed,
performing a hash-cons on each node.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/hashcons.png}
  \caption{Number of calls to \mintinline{haskell}{hashcons} to \mintinline{haskell}{add} operations performed} \label{fig:hashcons}
\end{figure}
As you can see from figure \ref{fig:hashcons}, this code will perform approximately
$2^{n+1}$ hashcons operations, where n is the number of {\bf add}'s.

\subsection{Explicit Sharing and Limitations} \label{limitexplicit}

@@

addChains x =
  let_ x (\x0 ->
  let_ (add x0 x0)  (\x1 ->
  let_ (add x1 x1)  (\x2 ->
   ...
  )))
\end{minted}
This makes the code a bit clunky and adds an extra burden on the DSL writer, but
it prevents unnecessary hash-consing in our example.

However the method has it's limitations, suppose we want to write a DSL function
that returns multiple
outputs, such as tuples or container types like lists (for example
\mintinline{haskell}{novel :: (repr Int,repr Int) -> (repr Int,repr Int)}). When
we go to implement DAG generation, we'll be forced to split the State monad into
two
\begin{minted}{haskell}
instance Exp Graph where
  ...
  novel e1 e2 = let
     g1 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    h3 <- hashcons $ Novel h1 h2
                    hashcons $ NovelOut1 h3)
     g2 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    h3 <- hashcons $ Novel h1 h2
                    hashcons $ NovelOut2 h3)
     in (g1,g2)
\end{minted}
Each output it returns will now have to be individually evaluated, so a
chain of DSL functions that output 2 or more values will suffer from the same
exponential scaling of hashcons operation. Furthermore, there's no good way to
explicitely define sharing via a custom let construct because of closure.

@@

instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (hashcons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (hashcons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST "nadd" [e1,e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> hashcons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST

@@

test x y = let
  z = cache "z" (add x y)
  in add z z
\end{minted}

The cache operation replaces the current byte string AST with a new label, and
we'll define a new operation \mintinline{haskell}{runCache} that will check if
the label already exists in the cache map before inserting it.
\begin{minted}{haskell}
runCache sAST mAddCache cacheMap = do
  case mAddCache of
    Nothing -> return ()
    Just sAST0 ->
      case Trie.lookup sAST cacheMap of
        Nothing -> let cacheMap' = insert sAST sAST0 cacheMap
                   in modify (\dag -> dag { dagCache = cacheMap' })
        Just sAST1 ->
            if sAST1 == sAST0
               then return ()
               else error $ "attempted to recache: " ++ show sAST

seqArgs inps =
  let
    seqArg (Graph sT sAST mCache) =
      do DAG trie cacheMap _ <- get
         runCache sAST mCache cacheMap
         ...
\end{minted}
We need to make sure we don't attempt to insert the same cache label for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the State monad to prevent or modify the subtitution. The best we can do is
crash the program, or if we use a monad tansformer, we ould use
Control.Monad.Except to through an exception. Either way it's up to the DSL
writer to insure they don't reuse the same label.

\section{Conclusion}
We have presented a method for constructing finally tagless style DSL's with
sharing detection, that allow for DSL functions to have multiple outputs via
built-in haskell tuples of container types. It also avoids the use of unsafe
referencing as performed when doing observable sharing like in the work of
\cite{gill:observablesharing}.

The method has its drawbacks in terms of memory usage, but these can be overcome
by explicitly specifying sharing. This does present an extra burden on the DSL
writer to implement explicit sharing when necessary and ensure labels are not
reused. Future work may investigate the use of a preprocessor or plugin to
automate explicit sharing.

\subsubsection{Acknowledgements} This work drew heavily upon the work of Oleg
Kiselyov, who's work on implicit and explicit sharing and finally tagless styles
are the basis of our method

@@

% LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless Trie
% LocalWords:  Kiselyov's haskell consing
% Local Variables:
% LaTeX-verbatim-environments-local: ("minted")
% eval: (setq-local LaTeX-indent-environment-list (cons '("minted" current-indentation) (default-value 'LaTeX-indent-environment-list)))
% End:
